---

Bike-share Analysis using R (presentation available in Kaggle notebook https://www.kaggle.com/hongnguyen999/bike-share-trip-analysis)

Data source: "https://divvy-tripdata.s3.amazonaws.com/index.html"

---

## Setting up environment 
Notes: setting up my R environment by loading the 'tidyverse', 'lubridate' and 'ggplot2' packages

```{r install packages}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
library("tidyverse")
library("lubridate")
library("ggplot2")
```
## Setting working directory to call data
```{r set up WD}
getwd()
setwd("C:/Users/An/Desktop/data analyst google/capstone project/bike_trips_case/R")
```

## Collecting data
Notes: upload bike_trips dataset

```{r collect data}
m12_2020 <- read.csv("202012-divvy-tripdata.csv")
m1_2021 <- read.csv("202101-divvy-tripdata.csv")
m2_2021 <- read.csv("202102-divvy-tripdata.csv")
m3_2021 <- read.csv("202103-divvy-tripdata.csv")
m4_2021 <- read.csv("202104-divvy-tripdata.csv")
m5_2021 <- read.csv("202105-divvy-tripdata.csv")
m6_2021 <- read.csv("202106-divvy-tripdata.csv")
m7_2021 <- read.csv("202107-divvy-tripdata.csv")
m8_2021 <- read.csv("202108-divvy-tripdata.csv")
m9_2021 <- read.csv("202109-divvy-tripdata.csv")
m10_2021 <- read.csv("202110-divvy-tripdata.csv")
m11_2021 <- read.csv("202111-divvy-tripdata.csv")
```
## Checking if column names of the files match
Notes: they have to match to be able to be joined into one file
```{r columns checking}
colnames(m12_2020)
colnames(m1_2021)
colnames(m2_2021)
colnames(m3_2021)
colnames(m4_2021)
colnames(m5_2021)
colnames(m6_2021)
colnames(m7_2021)
colnames(m8_2021)
colnames(m9_2021)
colnames(m10_2021)
colnames(m11_2021)
```
## Combining 12 individual data frames into one big data frame

```{r}
all_trips <- bind_rows(m12_2020,m1_2021,m2_2021,m3_2021,m4_2021,m5_2021,m6_2021,m7_2021,m8_2021,m9_2021,m10_2021,m11_2021)
```

## Removing unnecessary data fields that are not going to be used in our analysis: lat, long

```{r}
all_trips <- all_trips %>%  select(-c(start_lat,start_lng,end_lat,end_lng))
```

## Quick summary of how many observations fall under each type of customer: casual vs member

```{r}
table(all_trips$member_casual)
```

## Adding date, day, month, year columns. This will allow us to be able to aggregate ride data for each month, day, or year

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
```

## Adding another column to specify the day of the week for each ride

```{r}
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

## Add a "ride_length" calculation to "all_trips" (in seconds)

```{r}
all_trips$ride_length <-difftime(all_trips$ended_at,all_trips$started_at)
```

## Inspecting the structure of the columns
```{r}
str(all_trips)
```

## The "ride_length" is in numeric so that we can run the duration calculations

## Next, we remove bad data including trips with duration(ride_length) is equal to or less than 0. This will be a new verison of the dataframe (all_trips_v4)

```{r}
all_trips_v4 <- all_trips[all_trips$ride_length > 0,]
```

## Conducting descriptive analysis

```{r}
mean(all_trips_v4$ride_length)
median(all_trips_v4$ride_length)
max(all_trips_v4$ride_length)
min(all_trips_v4$ride_length)
```

## Now we conduct those calculations for each type of users; casual vs member

```{r}
aggregate(all_trips_v4$ride_length~all_trips_v4$member_casual,FUN=mean)
aggregate(all_trips_v4$ride_length~all_trips_v4$member_casual,FUN=median)
aggregate(all_trips_v4$ride_length~all_trips_v4$member_casual,FUN=max)
aggregate(all_trips_v4$ride_length~all_trips_v4$member_casual,FUN=min)
```

## Notice that the days of the week are out of order. Thus, we rearrange the days of the week into order.

```{r}
all_trips_v4$day_of_week <- ordered(all_trips_v4$day_of_week,levels=c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))
```

## Now we can see the average ride time for each day for casual and member customers with the days in order.

```{r}
aggregate(all_trips_v4$ride_length~all_trips_v4$member_casual+all_trips_v4$day_of_week,FUN=mean)

```

## Analyzing data by user type and weekday. Running calculations for number of rides and average duration.

```{r}
all_trips_v4 %>%
  group_by(member_casual,day_of_week)%>%
  summarise(number_of_rides=n(),average_duration=mean(ride_length))%>%
  arrange(member_casual,day_of_week)%>%
  print(width=Inf)
```
## Visualizing the number of rides by user type

```{r}
all_trips_v4 %>%
  group_by(member_casual,day_of_week)%>%
  summarise(number_of_rides=n(),average_duration=mean(ride_length))%>%
  arrange(member_casual,day_of_week)%>%
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Visualizing average duration by user type 

```{r}
all_trips_v4 %>%
  group_by(member_casual,day_of_week)%>%
  summarise(number_of_rides=n(),average_duration=mean(ride_length))%>%
  arrange(member_casual,day_of_week)%>%
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Export a summary file for further anlysis later 

```{r}
counts<-aggregate(all_trips_v4$ride_length~all_trips_v4$member_casual+all_trips_v4$day_of_week, FUN = mean)

```



